---
title: "Deliverable 2"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# CSV FIle
CSV File is to large to be hosted on github.  

```{r echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(rvest)
library(lubridate, warn.conflicts = FALSE)
library(weatherData)
library(caret)
library(modelr)
df <- read.csv("DelayedFlights.csv")
weather_data <- read.csv("San_Jose_2008_weather.csv")
```

# Introduction:
  My choice in data has to do with flight delays. I have personally been on many flights that alwasy seem to have some sort of delay. One of the craziest delays pertained to a simple light that had gone out which caused us to completely de-plane and took hours to find a new plane.   


# Data Set Info:
  This dataset is on the flight delays in the year 2008. This dataset includes the exact departure times along with the schedule departure times. There is also listings of delays or cancellations allowing me to analyze the average departure time per airport.   
  
# Data Set Variables/Units
**- Year:** 2008  (Integer)  
**- Month:** 1 - 12  (Integer)  
**- DayofMonth:** 1 - x (x = number of days in month)  (Integer)    
**- DayofWeek:** 1 - 7 (1 = Monday, 7 = Sunday)  (Integer)   
**- DepTime:** Exact departure time in format HHMM.SS (H = hour, M = minutes, S = seconds)  (Double)  
**- CRSDepTime:** The scheduled Arrival time in format HHMM (H = hour, M = minutes)  (Integer)  
**- ArrTime:** The exact Arrival time in format HHMM.SS (H = hour, M = minutes, S = seconds)  (Double)  
**- CRSArrTime:** The scheduled Arrival time in format HHMM (H = hour, M = minutes)  (Integer)  
**- UniqueCarrier:** The Carrier code for flight company. (EX AA = American Airlines)  (Factor)  
**- FlightNum:** Flight Number  (Integer)  
**- TailNum:** Tail Number  (Factor)  
**- ActualElapsedTime:** Exact Flight time in minutes (Double)     
**- CRSElapsedTime:** The scheduled or estimated flight time in minutes (Double)    
**- AirTime:** Time in air in minutes  (Double)  
**- ArrDelay:** The arrival delay in minutes.  (Double)  
**- DepDelay:** The Departure delay in minutes.  (Double)  
**- Origin:** IATA code for origin airport  (Factor)  
**- Dest:** IATA code for destination airport  (Factor)  
**- Distance:** Flight distance in miles  (Integer)  
**- TaxiIn:** Time to taxi to gate. Represented in minutes  (Double)  
**- TaxiOut:** Time to taxin to runway. Represented in minutes  (Double)  
**- Cancelled:** 0 = NO 1 = YES  (Integer)  
**- CancellationCode:** Cancellation Reason A = Carrier, B = weather, C = NAS(National Airspace System), D = security (Factor)  
**- Diverted:** 0 = NO, 1 = YES  (Integer)  
**- CarrierDelay:** Represented in minutes (Delay ex: bird strike, baggage loading)  (Double)  
**- WeatherDelay:** Represented in minutes (Delay ex: snow storm, de-icing)  (Double)  
**- NASDelay:** Represented in minutes (Delay ex: Airport operations, heavy traffic)  (Double)  
**- SecurityDelay:** Represented in minutes (Delay ex:Passenger screening)  (Double)  
**- LateAircraftDelay:** Represented in minutes (Delay ex: head wind)  (Double)  

## Definitions:
  **- NAS: ** This stands for the National Airspace System. This system includes every airport, services, airspace rules, personell, procedures, and equipment. This sysem also includes the military. Reasons for this system causing a delay would most likely be due to a military procedure like training in an area casuing a delay to the flight path to avoid the training. 
  
# Data Science Questions Trying to be answered:  
6. Can we predict and shorten delay times by changing airlines at specific airports.


## Arrival Delay vs Departure Delay per airport
The below graph shows the Arrival Delay and departure delays with each dot representing a different airport.  
One thing to notice is the dots around the 2500 mark. Being that these numbers are representing minutes delayed this is rather High at that specific airport. As you can see from the summary it seems that the airport in question is MSP or the Minneapolis Saint Paul International airport. This delay roughly translates to a 1.7 day delay for the flight. Another thing to note as displayed below is that the delay is due to NAS.     

```{r}
 ggplot(data = df, aes(x = DepDelay, y = ArrDelay)) +
  geom_point(aes(color = Dest), show.legend = FALSE, na.rm = TRUE) +
  labs(title = "Airports Arrival and Departure delays", x = "Departure Delays (min)", y = "Arrival Delays (min)")

print(filter(df, DepDelay > 2000))
```

# Sumary of Delays data
Looking at the summarry below some interesting things to note are the negative values shown. At first glance this seems incorrect but after looking more indepth these negative values in the delay columns are simply saying that the airplane is actually ahead of schedule either due to every passenger boarding early, or something simple like a tailwind during the flight.
```{r}
summary(df$ArrDelay)
summary(df$DepDelay)
```


# Cancelled flights Per Origin Airport
This is showing the number of cancelled flights per carrier colord by the cancellation code. Notice there wasnt a single security cancellation this can most likely be due to the NAS cancellations being almost the same thing. Also notice that the Carrier cancellations were quite high for United Airlines(UA), American Airlines(AA), Skywest (OO), and Continental Airlines(CO). Notice also that there is a significant number of cancelled flight with the weather cancellation code for Skywest(OO) and Envoy Airlines(MQ).
```{r}
 ggplot(data = filter(df, CancellationCode!= "N"), aes(x = UniqueCarrier )) +
  geom_bar(aes(fill = CancellationCode), show.legend = TRUE, position = "dodge") +
   labs(title = "Cancelled Flights", x = "Airline Iata Code", y = "Number of cancelled flights") +
  scale_fill_manual(labels = c("Carrier Cancellation", "Weather Cancellation", "NAS Cancellation"), values = c("Red", "Blue", "Green"))
```

# Carrier Delays Per Airline THROWOUT
As you can see below are the Carrier Cause delays. The Delays are all colored by Origin airport suggesting that most of these delays have to do with either some sort of flight clearance or gate clearance. One interesting thing to lookat is how Northwest airlines seems to have a rather high carrier delay with American Airlines and American Eagle airlines following closly behind.

```{r}
ggplot(data = df, aes(x = UniqueCarrier, y = CarrierDelay)) +
  geom_point(aes(color = Origin), show.legend = FALSE, na.rm = TRUE)
```


# Summary Departure Delays
This simply summarizes the Departure Delays data showing the minimum departure delay for the entire year was 6 minutes. This is interesting as not a single flight departed on time showing some sort of inefficiency.     
CRSDepTime is the estimated departure time. 
```{r}
selectedCol <- c(10, 7, 6, 17, 18)
summary(df[selectedCol])

```

# Summary Arrival Delays
This data summarization is for the arrival delays. Looking closly something to notice is that the arrival delays has a minimum of -109 minutes. This is saying that in the best conditions with a very minimal departure delay the flight actually arrived 109 minutes early. This can be due to a tail wind or the pilots pushing to plane faster. 
```{r}
selectedCol <- c(10, 8, 9, 16, 19)
summary(df[selectedCol])
```



# DELIVERABLE 2
#make clean sub datascience question relating to weather


#in background originally i wanted to acomplish but ended up trying to focus on this. these would be my future steps

For this Deliverable my main goals were to generate a new dataset using a webscrape method. I was then to take the data and create a model to work towards my Data Science questions where i try and predict or shorten delays at airports. Originally my goals was to try and look at specific airlines and shorten delyas based on different airline providers. This did not seem to work so easily so instead what my goal is to look at the weather data for the year 2008 and compare it to the weather delays in 2008. This will give me the ability to see the delays based on weather at the SanJose Airport to see why there are weather delays. From this data I can see if its a specific airline that is being delayed or a group of airlines.  
Some of the limitations I have with this data is specifically the dirty data caused by possible electrical interference in the instruments.  

# Webscrape
Web Scrape Web Data. Due to web server authentication settings the data will be loaded from csv file below as web server causes a time out during web scraping sometimes. This data is for the San Jose International Airport Weather.  

```{r eval = FALSE}
data_date <- ymd(20080101)
weather_data <- tibble(Year = integer(), Month = integer(), Day = integer(), Minimum_Temp = character(), Maximum_Temp = character(), Total_Rain = character(), Visibility = character(), Max_Wind_Speed = character(), Max_Wind_Gust = character())
for(month in 1:5)
{
    for(i in 1:73) 
  {
    df1 <- read_html(paste("https://www.almanac.com/weather/history/zipcode/95110/", data_date, sep = "")) %>%
    html_table()
    
    df <- as_tibble(df1[[2]], .name_repair = "minimal")
    weather_data <- add_row(weather_data, Year = year(data_date), Month = month(data_date), Day = day(data_date), Minimum_Temp = df[1,2], Maximum_Temp = df[3,2], Total_Rain = df[8,2], Visibility = df[9,2], Max_Wind_Speed = df[13,2], Max_Wind_Gust = df[14,2])
    data_date <- data_date + 1
    Sys.sleep(5)
    closeAllConnections()
    gc()
    }
  
  Sys.sleep(60)
  gc()
}
weather_data<-data.frame(lapply(weather_data, as.character), stringsAsFactors=FALSE)

write.csv(as_data_frame(weather_data),"/Users/brettflorek/Portfolio/San_Jose_2008_weather.csv", row.names = FALSE)

```

# Load Pre-Collected data 
This will load the pre-collected webscrape data. This data was pre scraped due to the length of time taking to scrape the data and the webserver causing timeouts for too many connections form one source.  

```{r}

weather_data <- read.csv("San_Jose_2008_weather.csv")
print(weather_data)

```

# Data Set Variables
List of all the variables of the San Jose International Airport Weather Data Set  
**- Year:** 2008  (Integer)  
**- Month:** Month 1-12 (Integer)  
**- Day:** Day in month 1-31 respectively (Ineteger)  
**- Minimum_Temp** Minimum temperature for that day in Farenheit (Factor)  
**- Maximum_Temp** Maximum temperature for that day in Farenheit (Factor)  
**- Total_Rain** Total amount of rain for that day in Inches (Factor)  
**- Visibility** The visibility distance in miles (Factor)  
**- Max_Wind_Speed** Maximum wind speed recorded that day in MPH (Factor)  
**- Max_Wind_Gust** Maximum Wind Gust in MPH. Wind Speeds of 999MPH are electrical interference. (Factor)  

# Preparing data

Converted all factors to doubles and excluded all 999 MPH wind gusts as these are mostly caused by interference. Converted interference values to NA.
```{r}
index <- c(1:365)
mutating <- weather_data
weather_data_doubles <- tibble(Year = integer(), Month = integer(), Day = integer(), Minimum_Temp = double(), Maximum_Temp = double(), Total_Rain = double(), Visibility = double(), Max_Wind_Speed = double(), Max_Wind_Gust = double())
for(i in index)
{
  min_temp_double <- as.double(substr(as.character(mutating$Minimum_Temp[i]), 1, nchar(as.character(mutating$Minimum_Temp[i])) - 2))
  max_temp_double <- as.double(substr(as.character(mutating$Maximum_Temp[i]), 1, nchar(as.character(mutating$Maximum_Temp[i])) - 2))
  total_rain_double <- as.double(substr(as.character(mutating$Total_Rain[i]), 1, nchar(as.character(mutating$Total_Rain[i])) - 2))
  visibility_double <- as.double(substr(as.character(mutating$Visibility[i]), 1, nchar(as.character(mutating$Visibility[i])) - 2))
  max_wind_speed_double <- as.double(substr(as.character(mutating$Max_Wind_Speed[i]), 1, nchar(as.character(mutating$Max_Wind_Speed[i])) - 3))
  max_wind_gust_double <- as.double(substr(as.character(mutating$Max_Wind_Gust[i]), 1, nchar(as.character(mutating$Max_Wind_Gust[i])) - 3))
  
  if(max_wind_gust_double == 999)
  {
    max_wind_gust_double <- NA
  }
  
  weather_data_doubles <- add_row(weather_data_doubles, Year = mutating$Year[i], Month = mutating$Month[i], Day = mutating$Day[i], Minimum_Temp = min_temp_double, Maximum_Temp = max_temp_double, Total_Rain = total_rain_double, Visibility = visibility_double, Max_Wind_Speed = max_wind_speed_double, Max_Wind_Gust = max_wind_gust_double)
}
print(weather_data_doubles)

ggplot(data = weather_data_doubles, aes(x = Total_Rain, y = Visibility)) +
  geom_point()
```

# Model
This is the initial model showing Visibility agains the Total_Rain and Minimum_Temp
The reason I chose total Rain and Minimum temp for my model is that they both would have the best relation with the visibility factor.  
```{r}
model <- lm(Visibility ~ Total_Rain + Minimum_Temp, data = weather_data_doubles)
summary(model)
```

# Data split for model testing and training

I split the data using the average split of 80% for training and 20% for testing.  
Once the Data was split the model was recreated using the training data.  
This new model allowed for predictions to be made using the testing dataset.  
I then added my predictions back to my dataset.
```{r}
#split data
train_rows <- as.vector(createDataPartition(weather_data_doubles$Visibility, p = 0.8, list = FALSE))
#get training rows
train <- weather_data_doubles[train_rows, ]

#get testing rows
test <- weather_data_doubles[-train_rows, ]

#recreate model with training rows
model <- lm(Visibility ~ Minimum_Temp , data = train)

#get predictions CHANGE SPLIT SET AND TRAIN USING VALIDATION SET
predictions <- predict(model, test)
predictions <- add_predictions(test, model)

#LOOK AT RESIDUALS
#GOODNESS OF FIT MEASURES (R2)

#Residual PLOT
#below 0 means under predicting over 0 means over preditcting
```


# Predictions Graph
Below is a graph showing my predictions generated using the testing set with my model.  
```{r}
ggplot(data = predictions, mapping = aes(x = Visibility, y = pred)) + 
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red")
```





